Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2023-11-17 paper/markdown/manuscript.Rmd
Compatibility with new clustering tools
Changes in code for compatibility with the latest version of `clustTools`. Minor graphical changes in figures. No effects on results.

To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to medoid clustering in respect to a broad range of psychometric scores (__Supplementary Table S2__). 
Three mental health clusters were identified in the training subset of the study cohort: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress) and named after their key mental characteristic (__Supplementary Figure S3__). 
Subsequently, the mental health cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
(training: `r signif(semi_clust$stats$variance[1], 2)`, test: `r signif(semi_clust$stats$variance[2], 2)`), 
comparable average silhouette statistics 
(training: `r signif(semi_clust$stats$mean_sil[1], 2)`, test: `r signif(semi_clust$stats$mean_sil[2], 2)`), 
comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).

>>>>

To explore symptoms of PTSD and further facets of mental health following mountain sport accidents, we subjected the participants to medoid clustering in respect to a broad range of psychometric scores (__Supplementary Table S2__). 
Three mental health clusters were identified in the training subset of the study cohort: 'neutral', 'PTG' (post-traumatic growth) and 'PTS' (post-traumatic stress) and named after their key mental characteristic (__Supplementary Figure S3__). 
Subsequently, the mental health cluster assignment could be validated in the test subset as evident from comparable fractions of explained clustering variance 
(training: `r signif(semi_clust$stats$frac_var[1], 2)`, test: `r signif(semi_clust$stats$frac_var[2], 2)`), 
comparable average silhouette statistics 
(training: `r signif(semi_clust$stats$sil_width[1], 2)`, test: `r signif(semi_clust$stats$sil_width[2], 2)`), 
comparable cluster sizes, good visual cluster separation and high similarity between the corresponding clusters in the training and test cohort subsets (__Supplementary Figures S4 - S6__).

### 2023-11-17 paper/markdown/supplementary_material.Rmd
Compatibility with new clustering tools
Changes in code for compatibility with the latest version of `clustTools`. Minor graphical changes in figures. No effects on results.

Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
training (variance = `r signif(semi_clust$stats$variance[1], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[1], 2)`) and 
test subset (variance = `r signif(semi_clust$stats$variance[2], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

>>>>

Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
training (variance = `r signif(semi_clust$stats$frac_var[1], 2)`, 
average silhouette = `r signif(semi_clust$stats$sil_width[1], 2)`) and 
test subset (variance = `r signif(semi_clust$stats$frac_var[2], 2)`, 
average silhouette = `r signif(semi_clust$stats$sil_width[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).
